A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}

### 2014-10-26 CC/writeup.Rnw
Add write-up to github. Also made some additional comments about data structure.

  \subsection{Vertex Aesthetics}
  In our geometry, the possible vertex aesthetics mimic the usual aesthetics in \texttt{ggplot2} for points.  The \texttt{vlabel} aesthetic is the name or number assigned to each vertex in order to uniquely identify it.  This aesthetic is most useful for smaller network objects where all vertex names can be printed on or near their corresponding vertices and still be readable by the human eye.  The \texttt{vfill} aesthetic can be an identity color (e.g. \texttt{color = "red"} outside of the \texttt{aes} statement) to change the fill color of all vertices, or it can take a factor variable which identifies each vertex as belonging to one of several groups, and it fills the vertices of the graph according to this grouping.  It can also fill the vertices according to the different values of  some numeric variable, such as degree.  The colors are set to the default color palette in \texttt{ggplot2}, and the palette used can be changed in the usual way within the \texttt{ggplot2} framework, by using \texttt{scale\_fill\_brewer}.  The \texttt{vcolor} aesthetic operates in the same way as the \texttt{vfill} aesthetic, but it changes the outline color of the vertex instead of the fill color of the vertex, which is in line with the \texttt{ggplot2} use of the \texttt{color} and \texttt{fill} aesthetics.  The \texttt{vsize} aesthetic can also be set to an identity value to increase or decrease the size of all vertices in the graph, or it can be set to the degree of the vertices in undirected graphs or the in-degree (or out-degree) of directed graphs.  \footnote{Sam's questions/notes: Is ``degree" going to be a column in our data frame? Or will we have to write something to detect degree size?} It can also be set to any other numeric variable associated with the vertices in the network.  The \texttt{vshape} aesthetic is for changing the shape of the vertices from the default circle to other shapes, like square or triangle.  It can also change the shape of the vertices according to some grouping variable.  Finally, the \texttt{valpha} aesthetic will change the vertex transparency to a set value, like 1/10, or to some numeric variable of interest that takes on the values $[0,1]$.

@@

\end{document}

>>>>

  \subsection{Vertex Aesthetics}
  In our geometry, the possible vertex aesthetics mimic the usual aesthetics in \texttt{ggplot2} for points.  The \texttt{vlabel} aesthetic is the name or number assigned to each vertex in order to uniquely identify it.  This aesthetic is most useful for smaller network objects where all vertex names can be printed on or near their corresponding vertices and still be readable by the human eye.  The \texttt{vfill} aesthetic can be an identity color (e.g. \texttt{color = "red"} outside of the \texttt{aes} statement) to change the fill color of all vertices, or it can take a factor variable which identifies each vertex as belonging to one of several groups, and it fills the vertices of the graph according to this grouping.  It can also fill the vertices according to the different values of  some numeric variable, such as degree.  The colors are set to the default color palette in \texttt{ggplot2}, and the palette used can be changed in the usual way within the \texttt{ggplot2} framework, by using \texttt{scale\_fill\_brewer}.  The \texttt{vcolor} aesthetic operates in the same way as the \texttt{vfill} aesthetic, but it changes the outline color of the vertex instead of the fill color of the vertex, which is in line with the \texttt{ggplot2} use of the \texttt{color} and \texttt{fill} aesthetics.  The \texttt{vsize} aesthetic can also be set to an identity value to increase or decrease the size of all vertices in the graph, or it can be set to the degree of the vertices in undirected graphs or the in-degree (or out-degree) of directed graphs.  \footnote{Sam's questions/notes: Is ``degree" going to be a column in our data frame? Or will we have to write something to detect degree size?} It an also be set to any other numeric variable associated with the vertices in the network.  The \texttt{vshape} aesthetic is for changing the shape of the vertices from the default circle to other shapes, like square or triangle.  It can also change the shape of the vertices according to some grouping variable.  Finally, the \texttt{valpha} aesthetic will change the vertex transparency to a set value, like 1/10, or to some numeric variable of interest that takes on the values $[0,1]$.

@@

\section{More Thoughts (26 Oct 14)}
I was thinking some more about the data structure necessary to create a network, and I think it might be possible to just use one data frame.  If we have a data frame (called \texttt{df} below) with ``from" and ``to" columns as the ``x" and ``y" columns, respectively, we could get the vertex names from that by just using an internal command like:

<<r1, eval=FALSE>>=
vnames <- unique(c(unique(df$x), unique(df$y)))
@

This could simply be the default for the \texttt{vlabel} aesthetic. Then, the \texttt{vlabel} aesthetic could really just be a logical one, which gets the label of \texttt{vnames} as above if \texttt{vlabel = TRUE} and gets nothing if \texttt{vlabel = FALSE}.  To handle the rest of the aesthetics with just one dataframe,  the user would have to use \texttt{merge} or something of the like to join the vertex characteristics to the edge data frame and characteristics.  An example of this with the blood data set is below:

<<r2, echo=FALSE>>=
all <- paste(sort(rep(c('A', "B", 'O', "AB"),2)), rep(c("+","-"),4), sep = '')
A <- all[1:2]
B <- all[5:6]
AB <- all[3:4]
O <- all[7:8]

On <- cbind(from = rep('O-',8), to = all, letter = 'O')
Op <- cbind(from = rep('O+',4), to = c(AB[1], A[1], B[1],O[1]), letter = 'O')
Bn <- cbind(from = rep('B-',4), to = c(AB, B), letter = 'B')
Bp <- cbind(from = rep('B+',2), to = c(AB[1], B[1]), letter = 'B')
An <- cbind(from = rep('A-',4), to = c(AB, A), letter = 'A')
Ap <- cbind(from = rep('A+',2), to = c(AB[1], A[1]), letter = 'A')
ABn <- cbind(from = rep('AB-',2), to = AB, letter = 'AB')
ABp <- cbind(from = 'AB+', to = 'AB+', letter = 'AB')

blood <- data.frame(rbind(ABn, ABp, An, Ap, Bn, Bp, On, Op))

blood.vertices <- unique(blood[,c(1,3)])
names(blood.vertices) <- c("type", "group")
blood.edges <- blood[,1:2]
@
<<r3, tidy=TRUE>>=
blood.vertices
head(blood.edges)

names(blood.vertices) <- c('from', 'group')
blood.df <- merge(blood.edges, blood.vertices, by = 'from')
head(blood.df)
@
After this is done, the \texttt{vcolor} aesthetic could simply be set to \texttt{vcolor = group} and there wouldn't be any need for a separate data frame to be required to pass to \texttt{geom\_ net}. Furthermore, all of the edge aesthesics could simply be set to the variables that are columns in the \texttt{df}, which were determined prior to merging the vertex information with the edge information.  For example, suppose we wanted to color the edges of the blood donation network by whether or not the blood in the ``from" vertex is going to the same or different blood group. i.e. the edge from O- to O+ is the same color as the edge from A- to A+, etc. but the edge from O- to AB+ is a different color. Then the data would have looked like:

<<r4, tidy=TRUE>>=
blood.edges$group_to <- c(rep('same',3),rep('diff',2),rep('same',2),'diff','same',rep('diff',2),rep('same',2),'diff','same',rep('diff',6),rep('same',2),rep('diff',3),'same')
head(blood.edges)
blood.df <- merge(blood.edges, blood.vertices, by = 'from')
head(blood.df)
@

Then, the \texttt{ecolor} aesthestic would just be set to \texttt{ecolor = group\_to}.
\end{document}
